---
title: "Reproducible Research Project 1"
author: "Jeffrey M. Hunter"
date: "4/16/2019"
output:
  html_document:
    df_print: paged
  pdf_document: default
  html_notebook: default
  pdf_notebook: default
---

## Table of Contents

[Introduction](#introduction)<br />
[Set Up Environment](#set-up-environment)<br />
[Questions](#questions)<br />
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [Code for reading in the dataset and/or processing the data](#code-for-reading-in-the-dataset-andor-processing-the-data)<br />
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [Histogram of the total number of steps taken each day](#histogram-of-the-total-number-of-steps-taken-each-day)<br />
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [Mean and median number of steps taken each day](#mean-and-median-number-of-steps-taken-each-day)<br />
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [Time series plot of the average number of steps taken](#time-series-plot-of-the-average-number-of-steps-taken)<br />
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [The 5-minute interval that, on average, contains the maximum number of steps](#the-5-minute-interval-that-on-average-contains-the-maximum-number-of-steps)<br />
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [Code to describe and show a strategy for imputing missing data](#code-to-describe-and-show-a-strategy-for-imputing-missing-data)<br />
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [Histogram of the total number of steps taken each day after missing values are imputed](#histogram-of-the-total-number-of-steps-taken-each-day-after-missing-values-are-imputed)<br />
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends](#panel-plot-comparing-the-average-number-of-steps-taken-per-5-minute-interval-across-weekdays-and-weekends)<br />
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [All of the R code needed to reproduce the results (numbers, plots, etc.) in the report](#all-of-the-r-code-needed-to-reproduce-the-results-numbers-plots-etc-in-the-report)

## Introduction

It is now possible to collect a large amount of data about personal movement
using activity monitoring devices such as a
[Fitbit](http://www.fitbit.com/),
[Nike Fuelband](http://www.nike.com/us/en_us/c/nikeplus-fuelband), or
[Jawbone Up](https://jawbone.com/up). These type of devices are part of the
"quantified self" movement - a group of enthusiasts who take measurements about
themselves regularly to improve their health, to find patterns in their
behavior, or because they are tech geeks. But these data remain under-utilized
both because the raw data are hard to obtain and there is a lack of statistical
methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device.
This device collects data at 5 minute intervals through out the day. The data
consists of two months of data from an anonymous individual collected during the
months of October and November, 2012 and include the number of steps taken in 5
minute intervals each day.

The data for this assignment can be downloaded from the course web site.

The variables included in this dataset are:

* **steps**: Number of steps taking in a 5-minute interval (missing values are
  coded as ```NA```)
* **date**: The date on which the measurement was taken in YYYY-MM-DD format
* **interval**: Identifier for the 5-minute interval in which measurement was
  taken

The dataset is stored in a comma-separated-value (CSV) file and there are a
total of 17,568 observations in this dataset.

## Set Up Environment

Disable warnings

```{r setup, include = TRUE}
knitr::opts_chunk$set(echo = FALSE)
```

Load libraries

```{r, echo = TRUE}
library(dplyr, warn.conflicts = FALSE)
library(ggplot2)
```

## Questions

Reports will be run to answer specified in the homework assignment.

### Code for reading in the dataset and/or processing the data

**Load and Process the Dataset**

```{r, echo = TRUE}
setwd("~jhunter/repos/coursera/reproducible-research-course-project-1")
activityDataFile <- "data/activity.csv"
if (!file.exists(activityDataFile)) {
    tempFile <- tempfile()
    download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", tempFile)
    unzip(tempFile, exdir = "data")
    unlink(tempFile)
}
activityData <- read.csv(activityDataFile, sep = ",")
activityData$date <- as.POSIXct(activityData$date, format = "%Y-%m-%d", tz = "")
weekDay <- weekdays(activityData$date, abbreviate = FALSE)
activityData <- cbind(activityData, weekDay)
names(activityData)[4] <- "week_day"
```

**Display Dataset Summary**

```{r, echo = TRUE}
names(activityData)
```

```{r, echo = TRUE}
str(activityData)
```

```{r, echo = TRUE}
summary(activityData)
```

```{r, echo = TRUE}
head(activityData)
```

**Display Number of Missing Data Values**
```{r, echo = TRUE}
sum(is.na(activityData$steps) == TRUE)
```

### Histogram of the total number of steps taken each day

1. **Calculate Steps per Day (excluding missing values)**
    
    ```{r, echo = TRUE}
    stepsPerDay <- with(activityData, aggregate(steps, list(date), FUN = sum, na.rm = TRUE))
    colnames(stepsPerDay) <- c("Date", "Steps")
    head(stepsPerDay)
    ```

1. **Display Histogram**
    
    ```{r, echo = TRUE}
    gp1 <- ggplot(stepsPerDay, aes(Steps))
    gp1 + geom_histogram(boundary = 0, binwidth = 2500, col = "darkblue", fill = "lightblue") + ggtitle("Histogram of Steps per Day") + xlab("Steps") + ylab("Frequency") + theme(plot.title = element_text(face = "bold", size = 12, hjust = 0.5)) + scale_x_continuous(breaks = seq(0, 25000, 2500)) + scale_y_continuous(breaks = seq(0, 18, 2))
    ```

### Mean and median number of steps taken each day

1. **Mean (excluding missing values)**
    
    ```{r, echo = TRUE}
    mean(stepsPerDay$Steps, na.rm = TRUE)
    ```

1. **Median (excluding missing values)**
    
    ```{r, echo = TRUE}
    median(stepsPerDay$Steps, na.rm = TRUE)
    ```

### Time series plot of the average number of steps taken

Calculate steps per time interval
    
```{r, echo = TRUE}
stepsPerDayAvg <- aggregate(steps ~ interval, data = activityData, FUN = mean, na.action = na.omit)
stepsPerDayAvg$time <- stepsPerDayAvg$interval/100
colnames(stepsPerDayAvg) <- c("Time", "Steps")
```

Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r, echo = TRUE}
gp2 <- ggplot(stepsPerDayAvg, aes(Time, Steps))
gp2 + geom_line(col="darkblue") + ggtitle("Average Steps per Time Interval") + xlab("Time") + ylab("Steps") + theme(plot.title = element_text(face = "bold", size = 12, hjust = 0.5))
```
    
### The 5-minute interval that, on average, contains the maximum number of steps

Which 5-minute interval, on average across all the days in the dataset, contains
the maximum number of steps?

```{r, echo = TRUE}
activityData %>% group_by(interval) %>% 
    summarize(mean_by_interval = mean(steps, na.rm = TRUE)) %>%
    filter(mean_by_interval == max(mean_by_interval))
```

### Code to describe and show a strategy for imputing missing data

*Pending*

### Histogram of the total number of steps taken each day after missing values are imputed

*Pending*

### Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends

*Pending*

### All of the R code needed to reproduce the results (numbers, plots, etc.) in the report

*Pending*
